What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines

### 2023-08-28 paper/main.tex
fixes through beginning of page 4

\begin{definition}[$\omega$-regular expressions]\label{def:omega-regex}
    Our definition \ref{def:regex} of regular expressions concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model them using infinite regular expressions (i.e., $\omega$-regular expressions), which we define via the following grammar:
    \begin{align*}
        B = A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expressions]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length strings over fixed alphabet $\Sigma$. Let $\L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then $\L_{\omega}(B)$ is inductively defined as
    \begin{align*}
        \L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in \L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in \L(A)$} \\
        \L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in \L(A) \text{ and } s_2 \in \L_{\omega}(B)\} \\
        \L_{\omega}(B_1 + B_2) &= \L_{\omega}(B_1) \cup \L_{\omega}(B_2)
    \end{align*}
\end{definition}

@@

\subsection{Timeline}\label{sec:timeline}
We present timelines as graphic visualizations. It contains the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start''.
    \item Every node represents one time step, and each node has a propositional logic formula $\psi$, which specifies the behavior of atomic propositions at that time step. The formula $\psi$ must be true at that time step. If $\psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item A node in egg shape with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting points.
    \item Timelines can be in parallel, signifying that either of the parallel timeline could happen.
\end{itemize}

The definition of timeline we use here are specifically what we devised for the purpose of visualizing \ltl\ formulas. We show how to construct such timelines from \ltl\ formulas in \cref{sec:correctness-timeline}.

@@

    In \cref{fig:ex2}, one can reason about two parallel timelines: the upper timeline starts with $p$ at the first time step, followed by entering the grey box with one step of $\lnot p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $\lnot p$, and so on. The lower timeline starts with one step of $\lnot p$ and one step of $p$ outside the grey box, and then enter the infinite run of $\lnot p$ and $p$ repeating.

@@

        \caption{Example timeline for specification ``$p$ oscillates every time step''}

@@

    In \cref{fig:ex15}, one can reason about one timeline: the atomic proposition $a$ is false for finitely many time steps as signified by the second node (could also be zero time step); followed by one node with $a$ which substantiates the specification of ``$a$ eventually true''; once $a$ is true at some point, the later time steps can behave arbitrarily as signified by the infinite run of $\top$ (true) in the grey box.

@@

        \caption{Example timeline for specification ``$a$ is eventually true''}

@@

\section{Algorithm: ``Though she be but little, she is fierce$^*$''}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{A Midsummer Night's Dream}}
\endgroup
On a high level, our algorithm of converting \ltl\ formulas into timeline visualizations (depicted in \Cref{fig:algo}) works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure*}

\subsection{LTL to BA} \label{ltl2aut}
The automata-theoretic approach~\cite{ORBi-a8d60ab6-1101-4434-9511-c01ea4e5a15b} of studying \ltl\ (i.e., reducing \ltl\ formulas to their corresponding \Buchi automata) has been well-studied. Many algorithms and optimizations have been studied such as~\cite{DGV99, F03}. Our tool \tool uses SPOT~\cite{Dur22}\footnote{For our tool, we use SPOT version 2.11.} for this step, which provides an implementation of the translation from \ltl\ formulas to \Buchi automata.

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_\mathrm{sf}$), and for those for paths looping from the final state back to itself (say $r_\mathrm{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_\mathrm{sf}r_\mathrm{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in the digraph of the automaton.

The algorithm is outlined below (see Algorithms 1 - 4). %\cref{alg:ba2wregex}

@@

%\section{Theoretical Analyses} \label{sec:analysis}
\section{``If we are true to ourselves, we can not be false to anyone$^*$''}
\label{sec:analysis}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{Hamlet}}
\endgroup

@@

%\section{Tool showcase}
\section{``Be great in act, as you have been in thought$^*$''}
\label{sec:tool-showcase}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{King John}}
\endgroup
To demonstrate the result of our tool, we present three examples, one from a real-world model checking example from Zhao and Rozier~\cite{ZR14}, one from a randomly generated \ltl\ formula, and one specifically demonstrates the use of our tool in specification verification.

@@

\section{Thought is free$^*$: Artifact availability}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{The Tempest}}
\endgroup
The tool we present in this paper is available at
\[
    \texttt{\url{https://github.com/EULIR/ltl-explainability}}
\]
which comes with two command-line tools, \texttt{ltl2regex} and \tool. The specific usage of the tool can be found  \href{https://github.com/EULIR/ltl-explainability#usage}{here}.

>>>>

\begin{definition}[$\omega$-regular expressions]\label{def:omega-regex}
    Our definition \ref{def:regex} of regular expressions concerns only finite-length strings. However, since \ltl\ formulas reason about events that happen over an infinite-length timeline, we need to model them using infinite regular expressions (i.e., $\omega$-regular expressions), which we define via the following grammar:
    \begin{align*}
        B = A^{\omega} \mid AB \mid B + B
    \end{align*}
\end{definition}
\begin{definition}[Semantics of $\omega$-regular expressions]\label{def:omega-semantics}
    Let $\Sigma^{\omega}$ denote the set of infinite-length strings over fixed alphabet $\Sigma$. Let $\L_{\omega}(B)$ denote the $\omega$-language accepted by $\omega$-regular expression $B$. Then we define $\L_{\omega}(B)$ inductively as:
    \begin{align*}
        \L_{\omega}(A^{\omega}) & = \{s_1s_2s_3\cdots \mid s_i \in \L(A) \text{ and } i \ge 1\} \tag{$\epsilon
         \not\in \L(A)$} \\
        \L_{\omega}(AB) & = \{s_1s_2 \mid s_1 \in \L(A) \text{ and } s_2 \in \L_{\omega}(B)\} \\
        \L_{\omega}(B_1 + B_2) &= \L_{\omega}(B_1) \cup \L_{\omega}(B_2)
    \end{align*}
\end{definition}

@@

\subsection{Timelines}\label{sec:timeline}
We present timelines as graphic visualizations containing the following features:
\begin{itemize}
    \item Every timeline starts with a node named ``start.''
    \item Every node represents one time step, and each node has a propositional logic formula $\psi$, which specifies the behavior of atomic propositions at that time step. The formula $\psi$ must be true at that time step. If $\psi = 1$, that means all atomic propositions can behave arbitrarily.
    \item An egg-shaped node with a caption ``repeats $0$ - $\infty$'' means to repeat the (one time step) current node for arbitrarily finitely many times.
    \item A node with label ``$\cdots$'' means to repeat the pattern prior to it and after it arbitrarily finitely many times.
    \item The grey box means repeat infinitely. Once we reach the end of a timeline in the grey box, we must reenter the same grey box from any of its (left) starting points.
    \item Timelines can occur in parallel, signifying that any of the parallel timelines could happen.
\end{itemize}

The manifestation of timelines we use here is one of possibly many timeline representations that we devised for the purpose of visualizing \ltl\ formulas. We show how to construct such timelines from \ltl\ formulas in \cref{sec:correctness-timeline}.

@@

    In \cref{fig:ex2}, one can reason about two parallel timelines: the upper timeline starts with $p$ holding in the first time step, followed by entering the grey box with one step of $\lnot p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $\lnot p$, and so on. The lower timeline starts with one step of $\lnot p$ and one step of $p$ outside the grey box, and then enters the infinite run of $\lnot p$ and $p$ repeating.

@@

        \caption{Example of one possible timeline for the specification ``$p$ oscillates every time step''}

@@

    In \cref{fig:ex15}, one can reason about one timeline: the atomic proposition $a$ is false for finitely many time steps as signified by the second node (note that this could be zero time steps); followed by one node with $a$ that substantiates the specification of ``$a$ is eventually true.'' Once $a$ is true at some point, the later time steps can behave arbitrarily as signified by the infinite run of $\top$ (true) in the grey box.

@@

        \caption{Example of a timeline for the specification ``$a$ is eventually true''}

@@

\section{Algorithm: ``Though she be but little, she is fierce\textsuperscript{\textsection}''}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{A Midsummer Night's Dream}}
\endgroup
On a high level, our algorithm for converting \ltl\ formulas into timeline visualizations (depicted in \Cref{fig:algo}) works as follows.
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automaton. % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automaton. % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression. (Note that this step represents a stylistic choice balancing size, complexity, and clarity.) % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression as a timeline according to its structure. %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth, scale=0.3]{img/algorithm_outline.pdf}
    \caption{\tool \ algorithm outline}
    \label{fig:algo}
\end{figure*}

\subsection{LTL to BA} \label{ltl2aut}
The automata-theoretic approach~\cite{ORBi-a8d60ab6-1101-4434-9511-c01ea4e5a15b} to evaluating \ltl\ (e.g., via reducing \ltl\ formulas to their corresponding \Buchi automata) has been well-studied \cite{DGV99,F03,RV10,RV11,Dur14}. Our tool \tool\ uses SPOT~\cite{Dur22}\footnote{We used SPOT version 2.11.} for this step, which provides a provably-correct implementation of the translation from \ltl\ formulas to minimally-sized \Buchi automata.

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_\mathrm{sf}$), and for those for paths looping from the final state back to itself (say $r_\mathrm{ff}$), and finally combining those to form the $\omega$-expression for satisfying runs $\left(\bigplus_{f\in F} r_\mathrm{sf}r_\mathrm{ff}^\omega\right)$. Our visualization algorithm finds regular expressions for finite paths by iteratively deleting interior nodes in the digraph of the automaton.

%KYR: FIX THIS!
The algorithm is outlined below (see Algorithms 1 - 4). %\cref{alg:ba2wregex}

@@

%\section{Theoretical Analyses} \label{sec:analysis}
\section{``If we are true to ourselves, we can not be false to anyone\textsuperscript{\textsection}''}
\label{sec:analysis}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{Hamlet}}
\endgroup

@@

%\section{Tool showcase}
\section{``Be great in act, as you have been in thought\textsuperscript{\textsection}''}
\label{sec:tool-showcase}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{King John}}
\endgroup
To demonstrate the result of our tool, we present three examples, one from a real-world model checking example from Zhao and Rozier~\cite{ZR14}, one from a randomly generated \ltl\ formula, and one specifically demonstrates the use of our tool in specification verification.

@@

\section{Thought is free\textsuperscript{\textsection}: Artifact availability}
\begingroup\renewcommand\thefootnote{\textsection}
\footnotetext{William Shakespeare, \emph{The Tempest}}
\endgroup
The tool we present in this paper is available at
\[
    \texttt{\url{https://github.com/EULIR/ltl-explainability}}
\]
which comes with two command-line tools, \texttt{ltl2regex} and \tool. The specific usage of the tool can be found  \href{https://github.com/EULIR/ltl-explainability#usage}{here}.
